--- 
title: "Principal Component Analysis (PCA)"
author: "Payam Emami"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
link-citations: yes
description: "This is a short tutorial on how PCA works."
---

# Prerequisites

In order to run the code in this chapter, you will need to install a number of packages. The packages are listed below. The recommended way of installing the packages is through BioManager. 

The **bookdown** package can be installed from CRAN or Github:


```{r eval=TRUE,each=FALSE,message=FALSE,warning=FALSE,error=FALSE}

# Install packages
install.packages("bookdown")
install.packages("scatterplot3d")
install.packages("plotrix")
install.packages("formatR")
install.packages("heplots")

# load packages
library(scatterplot3d)
library(plotrix)


```

We will also use a dataset in this chapter. The dataset contains several thousand genes measured on 23 samples. The samples are coming from four groups and have been measured in two batches.

```{r}
# read the raw data
data<-read.table("data/b1_b2_data.gct",sep = "\t",header = T,comment.char = "#")

# remove unused columns 
data<-data[,-c(1:2)]

# transpose the data so that genes are in column and samples are in rows
data<-t(data)

# read the metadata
metadata<-read.table("data/b1_b2_sampleinfofile.txt",sep = "\t",header = T,comment.char = "#")


```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


